---
title: "R Users Group Week 4"
output:
  html_document: default
  html_notebook: default
---

# More Fun with Data Frames

We continue learning about data frames.  We start with examples of querying a data base to calculate population statistics.  Then we show an example of working with multiple data frames as a relational data base, illustrate the merge command and show how it can be useful in common tasks.  Our homework for this week is based on an actual ASRS data base of private markets investments.

# Logic Variables

First we need to learn some more R to have the tools needed to search and query databases.  Logical values are data type in R.  Here are some examples that illustrate this including truth tables for basic logic functions.  

```{r}
tf=c(TRUE,FALSE) # All caps TRUE and FALSE invoke logic values
str(tf)
```

```{r}
sum(tf) #summing on a logic variable returns the count of TRUE values
```

```{r}
vec=c('this is true','this is false')
vec[tf]  #indexing with a logic vector returns the true values
```

Here are truth tables for the basic logic functions.

```{r}
names(tf)=c('true','false')
outer(tf,tf,'==') #logical equals
```

```{r}
!tf   # not
```

```{r}
outer(tf,tf,'&')  # and
```

```{r}
outer(tf,tf,'|')  # or
```

When dealing with quantities, standard inequality expressions return logical results.  Here is an example that combines quantities, factors and logic operations.

```{r}
ages=c(20,23,6,50)
sex=c("F","M","F","F")
name=c("Sally","Tom","Betty","Lois")
sex=="M"  # logical vector showing who is male
which(sex=="M") # index of the males
ages[sex=="F"] # ages of the females
sum(sex=="M")  # how many men
sum(ages>23) #how many older than 23
sum(ages>=23) #how many 23 or older
sum((ages>21)&(sex=="M"))  #how many older than 21 and male
name[max(ages)==ages] #name of the oldest person
name[min(ages[sex=="F"])==ages]  #name of the youngest female
```

# Summarizing a book list with the table command

Here is a data frame of books that have won various awards and appeared on best books list.  Let's read it in.

```{r}
books=read.csv("books.csv")
dim(books)
colnames(books)
```

Let's answer some questions about the book list.


```{r}
authorcount=table(books$Author)  #count the number of books appearing on the list by each author on the list
authorcount=authorcount[order(authorcount,decreasing=TRUE)]  #sort the list so the authors with the largest number of books are on top
head(authorcount)
table(authorcount)  #now show how many authors by count of books on the list
read=books$read  #we are curious about which books have been read
head(read)  #but the data isn't formatted perfectly for our use
read[read==1]="read"  #cleaning it up by converting to a factor read/not read
read[is.na(read)]="not read"
x=table(books$Author,read)  #do the counting to create a table object
head(x)
str(x)
xdf=data.frame(Read=x[,which(colnames(x)=="read")],Books=rowSums(x))  #convert to a dataframe
xdf=xdf[order(xdf$Books,decreasing=TRUE),]  #sort with high count authors on top
head(xdf)  
```

# A relational database and the merge command

Build a implified database describing a hospital revenue context.  Populate with some random info using the sample command.

```{r}
patients=data.frame(patient=c("A","B","C","D","E"),insurance=c("UHC","Aetna","UHC","Aetna","UHC"))
patients
procedure_price=data.frame(procedure=c('a','b','c','d','a','b','c','d'),
                           insurance=c(rep("UHC",4),rep("Aetna",4)),
                           price=sample(10:20,8))
procedure_price
procedures_done=data.frame(patient=sample(patients$patient,20,replace=TRUE),
                           procedure=sample(procedure_price$procedure,20,replace=TRUE),
                           date=sample(c("Mon","Tue","Wed","Thu","Fri"),20,replace=TRUE))
procedures_done
```

These three dataframes constitue a relational database.  Note there are common columns in each.  Let's illustrate the merge command.

```{r}
procedures_done_ins=merge(procedures_done,patients)
procedures_done_ins
```

Note it has added the insurance company to the list of procedures done.  Now let's add the prices.

```{r}
hospital_revenue=merge(procedures_done_ins,procedure_price)
hospital_revenue
```

Now I have long data frame with all my procedures and prices.  I have what I need to analyze revenue.  Here we go.

```{r}
attach(hospital_revenue)  #allows me to refer to the database columns as variables -- saves typing but be careful
sum(price) #total revenue
aggregate(price,by=list(date),sum) #revenue by date
aggregate(price,by=list(patient),sum) #revenue by patient
aggregate(price,by=list(procedure),mean) #average revenue for each procedure
aggregate(price,by=list(insurance,procedure),sum) #revenue by insurance and procedure
```



# Homework

Answer some more questions about the book database.  Use the table command to find out how many books there are by score and then ask for each score how many have been read.  Have high score books been prioritized by the reader?

In the hospital example, add another dataframe to the database.  Suppose procedures a and b are by the oncology department and procedures c and d are done by pediatrics.  Create a dataframe that reflects that information, merge it, and analyze revenue by hospital department.

Read the fundinfo.csv file in the rug4 folder.  This is an actual ASRS file regarding private market investments.  It has columns showing investment names, portfolio assignments, portfolio categories, commitment amounts, unfunded amounts, etc.  Do some analysis on this dataframe -- how many investments are in each portfolio.  What is the commitment amount by portfolio.  Prepare a summary by portfolio of unfunded amounts in investments still in there investment periods (today() is less than expiration date of the investment period -- you need the lubridate package for this).
